Simulation Tools for Small Area Estimation: Introducing the \proglang{R}-package \proglang{saeSim}

### 2015-01-23 saeSim.tex
replies and changes referee 1

%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
%
%% include your article here, just as usual
%
% R CMD Sweave %.Rnw
% Loading (and installing) necessary packages.
%
\section{Introduction}
The demand for reliable small area statistics from sample surveys has been substantially grown over the last decades due to their use in public and private sectors. In this paper we present a framework for simulation studies inside the field of small area estimation. This tool might be useful for the prospective researcher or data analyst to provide reproducible research.

Reproducible research has become a widely discussed topic. In the field of statistics many open-source tools like the \proglang{R}-language \citep{r14} and \LaTeX, dynamic reporting packages like \proglang{knitr} \citep{yihui13}, \proglang{sweave} \citep{leisch02} and more recently \proglang{rmarkdown} \citep{allaire14}, make the integration of text and source code for statistical analysis possible. Publishing source code and data alongside research results draws special attention to authoring the analysis. However, the requirements for source code are different from the written words in the article itself.

@@

\section{Case studies}
\label{sec:caseStudy}
We present two applications of \proglang{saeSim}, one model-based simulation in Section \ref{sec:csModel} and a design-based simulation in Section \ref{sec:csDesign}. Before, we introduce some basic functionalities as the pipe operator (\proglang{\%>\%}) needs some explanation. The pipe operator is designed to make otherwise nested expressions more readable as a line can be read from left to right, instead from inside out \citep{bache14}. As a simple example see the following lines which are equivalent:

\begin{Schunk}
\begin{Sinput}
> sum(1:10)
> 1:10 %>% sum
\end{Sinput}
\end{Schunk}

In \proglang{saeSim} we rely on this operator, although all functions can be used without, we strongly recommend to use it. The following example shows some of the aspects of the package:

\begin{Schunk}
\begin{Sinput}
> setup1 <- sim_base_lm() %>% sim_sample(sample_number(5))
> setup2 <- sim_base_lm() %>% sim_sample(sample_fraction(0.05))
\end{Sinput}
\end{Schunk}

Without knowing anything about the setup defined in \proglang{sim\_base\_lm} we notice that \proglang{setup1} and \proglang{setup2} only differ in the applied sampling scheme. \proglang{sim\_sample} is responsible to control when a function is applied (after the population-level) and \proglang{sample\_number(5)} and \proglang{sample\_fraction(0.05)} define the explicit way of drawing samples. Separating the responsibility of each component into what is applied and when it is applied makes it possible to add new components to any step in the process. The composition of a simulation in that manner will focus on the definition of components and hide control structures. Any function can be passed to \proglang{sim\_sample} which has a \proglang{data.frame} as input as well as return value. The only responsibility of that function is to draw a sample, which makes it easy to find, understand and reuse when published. The operator \proglang{\%>\%} is used to add new components to the setup.

\subsection{Model-based simulation}
\label{sec:csModel}
In the following we show one way how to construct a simulation in a model-based setting. The aim is to estimate the domain predictions under a FH model. Involved components are \textit{data generation} and \textit{computing on aggregated data} (cf. Figure \ref{fig:flowdiagram}). The first step is to generate the data under the model:

@@

In this case the \textit{base-component} is a data table with an id variable named \proglang{idD} and constructed with the function \proglang{base\_id}. Any random number generator in \proglang{R} can be used. However, we have normally distributed variates, for which some predefined functions are available in the package.

\begin{Schunk}
\begin{Sinput}
> library(saeSim)
> setup <- base_id(nDomains = 40, nUnits = 1) %>% 
+   sim_gen_x(mean = 0, sd = 4) %>%
+   sim_gen_v(mean = 0, sd = 1)
> setup
\end{Sinput}
\begin{Soutput}
  idD          x          v
1   1 -2.5058152 -0.1645236
2   2  0.7345733 -0.2533617
3   3 -3.3425144  0.6969634
4   4  6.3811232  0.5566632
5   5  1.3180311 -0.6887557
6   6 -3.2818735 -0.7074952
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
> library(sae)
> comp_FH <- function(dat) {
+   modelFH <- eblupFH(y ~ x, vardir, data = dat)
+   dat$FH <- modelFH$eblup
+   dat
+ }
> setup <- setup %>% sim_comp_agg(comp_FH)
> setup
\end{Sinput}
\begin{Soutput}
  idD         x          v vardir          e         y  trueStat        FH
1   1  1.637607  0.7073107    0.1  0.1258998 104.10843 103.98253 104.06121
2   2  6.755493  1.0341077    0.2 -0.1822523 114.36284 114.54509 114.23876
3   3  6.346354  0.2234804    0.3  0.7253263 113.64151 112.91619 113.45213
4   4 -1.323631 -0.8787076    0.4 -0.4434978  96.03053  96.47403  96.45416
5   5 -9.140942  1.1629646    0.5 -0.4105563  82.47052  82.88108  82.46045
6   6  9.990646 -2.0001649    0.6 -0.7754272 117.20570 117.98113 118.08379
\end{Soutput}
\end{Schunk}

@@

An additional variable \proglang{idR} is automatically added as an ID-variable for the iteration as well as a variable \proglang{simName} to distinguish between scenarios. At this time we do not provide further tools to process the resulting data. Many packages are available in \proglang{R} to handle \proglang{data.frame}. In the design-based scenario we show how to process the result data into graphs with only a few lines of code.

@@

Using this data as population, we repeatedly draw samples from it. Then we predict the domain means by using a direct estimator and a unit-level model. The sampling design is to draw a 10 per cent sample from each region with simple random sampling. For each region the direct estimator for income and the EBLUP under the BHF model is computed. Although the data offers some more information, we only use \proglang{gender}, \proglang{age} and \proglang{agesq} as covariates. The function \proglang{eblupBHF} from the package \proglang{sae} is an implementation of the BHF estimator. This function expects three data objects and returns domain predictions. The three objects are the sampled data, the population means of the auxiliary variables and the population sizes in each domain. Before we begin to construct the simulation setup, we store these data tables as attributes to the population data. There are other options but adding attributes to the data is a very flexible form of processing data on different aggregation levels.

@@

In the next step, we define components which add the estimates of interest to the data. Here we compute the direct estimator of the mean income in each domain and the EBLUP under the BHF model. Although this could be done in one step, we separate the two computations to illustrate how to combine several estimations and how to define each component independent of each other. This focus on the definition of each component and  on meeting the convention of the defined interface is the intended approach. It automatically organises the simulation and each component is arranged using the simulation framework. Hence, we define two functions, one for adding the direct estimates and one for adding the EBLUP.

\begin{Schunk}
\begin{Sinput}
> comp_direct <- function(dat) {
+   attr(dat, "sampleMean") <- 
+     dat %>% group_by(region) %>% summarise(direct = mean(eqIncome))
+   dat
+ }
> comp_BHF <- function(dat) {
+   popMeans <- select(attr(dat, "popMeans"), -trueStat)
+   modelBHF <- 
+     eblupBHF(eqIncome ~ age + agesq + gender, region, meanxpop = popMeans,
+              popnsize = attr(dat, "popN"), data = dat)
+   attr(dat, "BHF") <- modelBHF$eblup
+   dat
+ }
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
> library(reshape2)
> ggDat <- melt(simResults,
+               id.vars = c("region", "trueStat"), 
+               measure.vars = c("BHF", "direct"), 
+               variable.name = "method",
+               value.name = "prediction")
\end{Sinput}
\end{Schunk}

@@

The package source is available on CRAN (\url{http://cran.r-project.org/web/packages/saeSim/}) and the repository for development on GitHub (\url{https://github.com/wahani/saeSim}). As GitHub allows to share and contribute source code using version-control, it is open for submissions. Apart from the availability of specific utility functions, we may promote and support the design of source code for simulation studies. One aspect is the design of simulations as processes of data. Furthermore, we encourage the definition of small and self contained components, i.e. functions. This reduces the lines of code necessary to be read in order to understand it's purpose.

The package provides more features than introduced in this article. To mention is the support of outlier contaminated data. At this time only representative outliers (outlying observations in the population) are supported \citep[cf.][]{Cha86}. However, we plan to extend this feature to non-representative outliers (outliers are part of the sample but not the population). Furthermore, an interface to \proglang{R} random number generators is available. The user can also generate group effects as needed in mixed models. As the response is created by an \proglang{R} expression, any form of non-linearity in the relationship between response and auxiliary variables as well as error components can be modelled.

>>>>

%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\red}[1]{\color{red} #1 \color{black} }

\begin{document}
%
%% include your article here, just as usual
%
% R CMD Sweave %.Rnw
% Loading (and installing) necessary packages.
%
\section{Introduction}
The demand for reliable small area statistics from sample surveys \red{has grown substantially} over the last decades due to their use in public and private sectors. In this paper we present a framework for simulation studies inside the field of small area estimation. This tool might be useful for the prospective researcher or data analyst to provide reproducible research.

Reproducible research has become a widely discussed topic. In the field of statistics many open source tools like the \proglang{R}-language \citep{r14} and \LaTeX, dynamic reporting packages like \proglang{knitr} \citep{yihui13}, \proglang{sweave} \citep{leisch02} and more recently \proglang{rmarkdown} \citep{allaire14}, make the integration of text and source code for statistical analysis possible. Publishing source code and data alongside research results draws special attention to authoring the analysis. However, the requirements for source code are different from the written words in the article itself.

@@

\section{Case studies}
\label{sec:caseStudy}
We present two applications of \proglang{saeSim}, one model-based simulation in Section \ref{sec:csModel} and a design-based simulation in Section \ref{sec:csDesign}. Before, we introduce some basic functionalities as the pipe operator (\proglang{\%>\%}) needs some explanation. The pipe operator is designed to make otherwise nested expressions more readable as a line can be read from left to right, instead from inside out \citep{bache14}. \red{As a simple example see the following lines which are equivalent with respect to their functionality:}

\begin{Schunk}
\begin{Sinput}
> library(magrittr)
> colMeans(matrix(rnorm(10), ncol = 2))
> rnorm(10) %>% matrix(ncol = 2) %>% colMeans
\end{Sinput}
\end{Schunk}

In \proglang{saeSim}, we rely on this operator. Although all functions can be used without it, we strongly recommend its usage. The following example shows some of the aspects of the package:

\begin{Schunk}
\begin{Sinput}
> library(saeSim)
> setup1 <- sim_base_lm() %>% sim_sample(sample_number(5))
> setup2 <- sim_base_lm() %>% sim_sample(sample_fraction(0.05))
\end{Sinput}
\end{Schunk}

Without knowing anything about the setup defined in \proglang{sim\_base\_lm} we notice that \proglang{setup1} and \proglang{setup2} only differ in the applied sampling scheme. \proglang{sim\_sample} is responsible to control when a function is applied (after the population-level) and \proglang{sample\_number(5)} and \proglang{sample\_fraction(0.05)} define the explicit way of drawing samples. Separating the responsibility of each component into what is applied and when it is applied makes it possible to add new components to any step in the process. The composition of a simulation in that manner will focus on the definition of components and hide control structures. Any function can be passed to \proglang{sim\_sample} which has a \proglang{data.frame} both as input and as return value. The only responsibility of that function is to draw a sample, which makes it easy to find, understand and reuse when published. The operator \proglang{\%>\%} is used to add new components to the setup.

\subsection{Model-based simulation}
\label{sec:csModel}

In the following we show one way how to construct a simulation in a model-based setting. The aim is to estimate the domain predictions under a FH model. Involved components are \textit{data generation} and \textit{computing on aggregated data} (cf. Figure \ref{fig:flowdiagram}). The first step is to generate the data under the model:

@@

In this case the \textit{base-component} is a data frame with an id variable named \proglang{idD} and constructed with the function \proglang{base\_id}. Any random number generator in \proglang{R} can be used. However, we have normally distributed variates, for which some predefined functions are available in the package. \red{For the reproducibility of the following results we also set the seed to 1. The seed is not part of a simulation setup in \proglang{saeSim} but needs to be supplied by the researcher.}

\begin{Schunk}
\begin{Sinput}
> set.seed(1)
> setup <- base_id(nDomains = 40, nUnits = 1) %>% 
+   sim_gen_x(mean = 0, sd = 4) %>%
+   sim_gen_v(mean = 0, sd = 1)
> setup
\end{Sinput}
\begin{Soutput}
  idD          x          v
1   1 -2.5058152 -0.1645236
2   2  0.7345733 -0.2533617
3   3 -3.3425144  0.6969634
4   4  6.3811232  0.5566632
5   5  1.3180311 -0.6887557
6   6 -3.2818735 -0.7074952
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
> comp_FH <- function(dat) {
+   modelFH <- sae::eblupFH(y ~ x, vardir, data = dat)
+   dat$FH <- modelFH$eblup
+   dat
+ }
> setup <- setup %>% sim_comp_agg(comp_FH)
> setup
\end{Sinput}
\begin{Soutput}
  idD         x          v vardir          e         y  trueStat        FH
1   1  1.637607  0.7073107    0.1  0.1258998 104.10843 103.98253 104.06121
2   2  6.755493  1.0341077    0.2 -0.1822523 114.36284 114.54509 114.23876
3   3  6.346354  0.2234804    0.3  0.7253263 113.64151 112.91619 113.45213
4   4 -1.323631 -0.8787076    0.4 -0.4434978  96.03053  96.47403  96.45416
5   5 -9.140942  1.1629646    0.5 -0.4105563  82.47052  82.88108  82.46045
6   6  9.990646 -2.0001649    0.6 -0.7754272 117.20570 117.98113 118.08379
\end{Soutput}
\end{Schunk}

@@

An additional variable \proglang{idR} is automatically added as an ID-variable for the iteration as well as a variable \proglang{simName} to distinguish between scenarios. At this time we do not provide further tools to process the resulting data as there are many tools readily available in \proglang{R}. In the design-based scenario we show how to process the result data into graphs with only a few lines of code.

@@

Using this data as population, we repeatedly draw samples from it. Then we predict the domain means by using a direct estimator and a unit-level model. The sampling design is to draw a 10 per cent sample from each region with simple random sampling. For each region the direct estimator for income and the EBLUP under the BHF model is computed. Although the data offers some more information, we only use \proglang{gender}, \proglang{age} and \proglang{agesq} as covariates. The function \proglang{eblupBHF} from the package \proglang{sae} is an implementation of the BHF estimator. This function expects three data objects and returns \red{the} domain predictions. The three objects are the sampled data, the population means of the auxiliary variables and the population sizes in each domain.

\red{Before we begin to construct the simulation setup, we store these data frames as attributes to the population data. However, extra care is needed because not all functions for manipulating data frames in \proglang{R} will preserve their attributes. Functions in \proglang{saeSim} all support this approach. Yet, in user defined functions this needs to be kept in mind by the author. Adding attributes to the data is a very flexible form of processing data on different aggregation levels and we believe this to be a reasonable trade-off. Of course other options can be applied within the possibilities of the \proglang{R} language.}

@@

\red{In the next step, we define components which add the estimates of interest to the data. Here we compute the direct estimator of the mean income in each domain and the EBLUP under the BHF model. Although this could be done in one step, we separate the two computations to illustrate how to combine several estimations and how to define each component independent of each other. This automatically organises the simulation and each component is arranged using the simulation framework. Hence, we define two functions, one for adding the direct estimates and one for adding the EBLUP.}

\begin{Schunk}
\begin{Sinput}
> comp_direct <- function(dat) {
+   attr(dat, "sampleMean") <- 
+     dat %>% group_by(region) %>% summarise(direct = mean(eqIncome))
+   dat
+ }
> comp_BHF <- function(dat) {
+   popMeans <- select(attr(dat, "popMeans"), -trueStat)
+   modelBHF <- 
+     sae::eblupBHF(eqIncome ~ age + agesq + gender, region, meanxpop = popMeans,
+              popnsize = attr(dat, "popN"), data = dat)
+   attr(dat, "BHF") <- modelBHF$eblup
+   dat
+ }
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
> ggDat <- reshape2::melt(simResults,
+               id.vars = c("region", "trueStat"), 
+               measure.vars = c("BHF", "direct"), 
+               variable.name = "method",
+               value.name = "prediction")
\end{Sinput}
\end{Schunk}

@@

The package source is available on CRAN (\url{http://CRAN.R-project.org/package=saeSim}) and the repository for development on GitHub (\url{https://github.com/wahani/saeSim}). As GitHub allows to share and contribute source code using version control, it is open for submissions. Apart from the availability of specific utility functions, we may promote and support the design of source code for simulation studies. One aspect is the design of simulations as processes of data. Furthermore, we encourage the definition of small and self contained components, i.e. functions. This reduces the lines of code necessary to be read in order to understand its purpose.

The package provides more features than introduced in this article. To mention is the support of outlier contaminated data. At this time only representative outliers (outlying observations in the population) are supported \citep[cf.][]{Cha86}. \red{However, we plan to extend this feature to non-representative outliers (outliers are part of the sample but not the population, e.g., incorrectly recorded values). Furthermore, an interface to random number generators in \proglang{R} is available.} The user can also generate group effects as needed in mixed models. As the response is created by an \proglang{R} expression, any form of non-linearity in the relationship between response and auxiliary variables as well as error components can be modelled.
